---
title: 'Ecological Fractal Net data - species richness statistics'
output:
  word_document: default
header-includes: \usepackage[singlelinecheck=false]{caption}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r setup}
library(dplyr)
library(e1071)   # For skewness() and kurtosis()
library(knitr)
library(moments)
library(readxl)

# Load helpers and configuration
source(file.path("..", "scripts", "helpers.R"))

project_root <- find_project_root()

# config contains the names of the required EFN locations and the details of the input files for each location
config <- load_project_config(project_root)

```

**Author:** Johanneke Siljee (UU, NL)

**Report** unpublished, exploratory analysis, not peer-reviewed, for internal use only.

## Purpose

This report explores the Ecological Fractal Net (EFN) data, focusing on species richness statistics .

# EFN species richness statistics

```{r load-EFN-data}
# -------------------------------------------------
#  Load ecological data
# -------------------------------------------------

# Get all xlsx files from the processed_data file path, and put them together in a dataframe, with the last column being the filename (without extension), and the first column the Plotcode from the sheet Species_count_Plotcode, and the second column the unique_Species_count from the same sheet.

base_dir <- file.path(config$paths$processed_data, "combined_results")

# read the xlsx file with species richness data
excel_file <- file.path(base_dir, "All_locations_1m_specrich.xlsx")
df <- read_xlsx(excel_file)

```

## Histogram

```{r histogram-plot}
# -------------------------------------------------
# Distribution of species richness with a histogram
hist(df$unique_Species_count, main = "Histogram of species richness", xlab = "Species Richness", ylab = "Frequency", breaks = 20)

```

# Density

```{r density-plot}

# Distribution of species richness with a density plot
plot(density(df$unique_Species_count, na.rm = TRUE), main = "Density Plot of Species Richness", xlab = "Species Richness", ylab = "Density")

```

# Summary statistics

```{r summary-stats}

# print Summary statistics of species richness
summary(df$unique_Species_count)

# Boxplot of species richness
boxplot(df$unique_Species_count, main = "Boxplot of Species Richness", ylab = "Species Richness")

```


Standard deviation: **`r sd(df$unique_Species_count)`**.

Are there negative values in species richness? **`r any(df$unique_Species_count < 0)`**


```{r alpha-stats-table, echo=FALSE, results='asis'}

# Recalculate stats here if not available globally
min_alpha <- min(df$unique_Species_count, na.rm = TRUE)
mean_alpha <- mean(df$unique_Species_count, na.rm = TRUE)
median_alpha <- median(df$unique_Species_count, na.rm = TRUE)
max_alpha <- max(df$unique_Species_count, na.rm = TRUE)
sd_alpha <- sd(df$unique_Species_count, na.rm = TRUE)
cv_alpha <- sd_alpha / mean_alpha

# Build the data frame
summary_stats <- data.frame(
  Statistic = c("Minimum", "Mean", "Median", "Maximum", 
                "SD", "CV (%)", "Skewness", "Kurtosis"),
  Value = c(
    min_alpha,
    round(mean_alpha, 2),
    round(median_alpha, 2),
    max_alpha,
    round(sd_alpha, 2),
    round(cv_alpha * 100, 2),
    round(skewness(df$unique_Species_count, na.rm = TRUE), 3),
    round(kurtosis(df$unique_Species_count, na.rm = TRUE), 3)
  )
)

# Render the table
kable(summary_stats, 
      caption = "Statistics for Unique Species Count",
      col.names = c("Metric", "Value"),
      digits = 2,
      booktabs = TRUE,     # Adds professional horizontal lines
      align = c("l", "r")        # Left-align labels, right-align values
)
```
