---
title: 'Species Richness Predictions'
output:
  word_document: default
header-includes: \usepackage[singlelinecheck=false]{caption}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r setup}
library(readr)
library(dplyr)
library(terra)
library(viridisLite)
library(ggplot2)
library(reshape2)
library(spdep)
source(file.path("..", "scripts", "helpers.R"))
project_root <- find_project_root()
config <- load_project_config(project_root)
```

**Author:** Johanneke Siljee (UU, NL)

**Report** unpublished, exploratory analysis, not peer-reviewed, for internal use only.

## Purpose

This report captures the analysis currently implemented for species-richness prediction maps across spatial scales.

```{r load-rasters}
# Load predicted rasters

prediction_filenames <- config$reporting$prediction_map_filenames
dutch_crs <- config$reporting$target_crs

richness_scale1 <- rast(file.path(config$paths$output_maps, prediction_filenames[["Q"]]))
richness_scale2 <- rast(file.path(config$paths$output_maps, prediction_filenames[["1m"]]))
richness_scale3 <- rast(file.path(config$paths$output_maps, prediction_filenames[["100m_mean"]]))
richness_scale4 <- rast(file.path(config$paths$output_maps, prediction_filenames[["100m_total"]]))

crs(richness_scale1) <- dutch_crs
crs(richness_scale2) <- dutch_crs
crs(richness_scale3) <- dutch_crs
crs(richness_scale4) <- dutch_crs

# Align extents and resolutions if necessary
richness_scale2 <- resample(richness_scale2, richness_scale1, method = "bilinear")
richness_scale3 <- resample(richness_scale3, richness_scale1, method = "bilinear")
richness_scale4 <- resample(richness_scale4, richness_scale1, method = "bilinear")

# Load raster datasets representing species richness at different scales
richness_stack <- c(richness_scale1, richness_scale2, richness_scale3, richness_scale4)
names(richness_stack) <- c("0.25m","1m","100mean","100total")

```

# Predicted species richness plots


```{r plot-predictions}  
# Plot the predicted species richness maps for each scale

# Get the absolute min and max across ALL layers at once
# This calculates min(min(layer1), min(layer2), ...) etc.
global_stats <- global(richness_stack, c("min", "max"), na.rm = TRUE)

# Extract values
zmin <- global_stats$min[1] # The very first column's min value
zmax <- global_stats$max[1] # The very first column's max value

# Check if they exist
if (!is.na(zmin) && !is.na(zmax)) {
  cols <- turbo(n = 100)
  
  plot(richness_stack, 
       main = names(richness_stack), 
       col = cols,
       zlim = c(zmin, zmax))
} else {
  warning("Could not calculate range (all data might be NA)")
}

```

# Prediction patterns analysis

```{r Analyse prediction patterns}

# Normalize the stacked raster to a 0-1 scale
richness_norm <- (richness_stack - min(richness_stack)) / (max(richness_stack) - min(richness_stack))

# ------------------------------------------------------------------
# Histograms and Density Plots
# ------------------------------------------------------------------
# Examine the distribution of species richness values at each scale
richness_df <- data.frame(
  Scale1 = values(richness_stack[[1]]),
  Scale2 = values(richness_stack[[2]]),
  Scale3 = values(richness_stack[[3]]),
  Scale4 = values(richness_stack[[4]])
)

richness_long <- melt(richness_df, variable.name = "Scale", value.name = "Richness")  # Reshape for ggplot

ggplot(richness_long, aes(x = Richness, fill = Scale)) +
  geom_density(alpha = 0.4) +
  labs(title = "Density of Predictions Across Scales") +
  theme_minimal()

# To reduce overplotting, we can sample a subset of the data for visualization
# Sample a subset of the data
set.seed(123)  # For reproducibility
subset_richness <- richness_long[sample(nrow(richness_long), size = 100000), ]

# Re-plot with the subset
ggplot(subset_richness, aes(x = Richness, fill = Scale)) +
  geom_density(alpha = 0.4) +
  labs(title = "Density of Predictions Across Scales, with subset sampling") +
  theme_minimal()


```

# Hotspot analysis

```{r hotspot-analysis}  
# ------------------------------------------------------------------
# Hotspot Analysis
# ------------------------------------------------------------------
# Identify areas of high or low richness (hotspots) at each scale and compare them:
# Extract values from the first layer of the raster stack
values_richness <- values(richness_stack[[1]])  # Extract raster values as a vector

# Compute the 90th percentile (exclude NA values)
q90 <- quantile(values_richness, 0.9, na.rm = TRUE)

# Create a logical raster of hotspots (areas above the 90th percentile)
hotspots <- richness_stack[[1]] > q90

# Plot the hotspots
plot(hotspots, col = c("lightgrey", "red"), legend = FALSE, main = "90th Percentile Hotspots")

# Identify hotpots occuring both in smallest and largest scale
hotspots_scale1 <- richness_stack[[1]] > quantile(values(richness_stack[[1]]), 0.9, na.rm = TRUE)
hotspots_scale4 <- richness_stack[[4]] > quantile(values(richness_stack[[4]]), 0.9, na.rm = TRUE)
common_hotspots <- hotspots_scale1 & hotspots_scale4

# Plot common hotspots, use colouring normal for ecological hotspots.
plot(common_hotspots, col = c("lightgrey", "red"), legend = FALSE, main = "Common hotspots in 0.25m² and 100m total scales")


# calculate how much of the hotspots in hotspots_scale1 are also in common_hotspots
hotspot_overlap <- sum(values(hotspots_scale1) & values(common_hotspots), na.rm = TRUE) /
  sum(values(hotspots_scale1), na.rm = TRUE)
# calculate how much of the hotspots in hotspots_scale4 are also in common_hotspots
hotspot_overlap2 <- sum(values(hotspots_scale4) & values(common_hotspots), na.rm = TRUE) /
  sum(values(hotspots_scale4), na.rm = TRUE)

```

Proportion of 0.25m² hotspots also in common hotspots: `r round(hotspot_overlap * 100, 2)`%

Proportion of 100m total hotspots also in common hotspots: `r round(hotspot_overlap2 * 100, 2)`%


# Spatial Cross-Scale Comparison
## Differences Across Scales
Compute the difference between richness predictions across scales (absolute or relative changes)

``` {r scale_differences}
diff_scale1 <- richness_stack[[2]] - richness_stack[[1]]

# Convert the terra raster to a data frame
raster_df <- as.data.frame(diff_scale1, xy = TRUE, na.rm = TRUE)

#rename column "1m" to "layer"
raster_df <- dplyr::rename(raster_df, layer = names(diff_scale1)[1])


# Plot using ggplot2
ggplot(raster_df, aes(x = x, y = y, fill = layer)) +
  geom_raster() +  # Use raster tiles for the plot
  scale_fill_gradient2(
    low = "red",     # Color for negative values
    mid = "white",   # Color for zero
    high = "blue",   # Color for positive values
    midpoint = 0,    # Set zero as the neutral midpoint
    name = "Difference"  # Legend title
  ) +
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Difference Between 1m² and 0.25m²",
    x = "Longitude",
    y = "Latitude"
  )

# Compute difference between 100m total and 0.25m2
diff_scale2 <- richness_stack[[4]] - richness_stack[[1]]

# Convert the terra raster to a data frame
raster_df <- as.data.frame(diff_scale2, xy = TRUE, na.rm = TRUE)

#rename column "1m" to "layer"
raster_df <- dplyr::rename(raster_df, layer = names(diff_scale2)[1])

# Plot using ggplot2
ggplot(raster_df, aes(x = x, y = y, fill = layer)) +
  geom_raster() +  # Use raster tiles for the plot
  scale_fill_gradient2(
    low = "red",     # Color for negative values
    mid = "white",   # Color for zero
    high = "blue",   # Color for positive values
    midpoint = 0,    # Set zero as the neutral midpoint
    name = "Difference"  # Legend title
  ) +
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Difference Between 100m total and 0.25m2",
    x = "Longitude",
    y = "Latitude"
  )
```

